---
title: "Validation: remlax against asreml, lme4, sommer and nlme"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation: remlax against asreml, lme4, sommer and nlme}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
library(remlax)
checks <- rx_validation_results("checks")
pairs <- rx_validation_results("pairs")
bench <- rx_validation_results("benchmark")
vers <- attr(checks, "versions")
a_python <- isTRUE(tryCatch(rx_python_check(quiet = TRUE)$ok, error = function(e) FALSE))
```

This vignette answers one question: does remlax compute the same restricted
likelihood, and reach the same optimum, as the programs a reader already
trusts? Every comparison below is the same model fitted to the same data by
remlax and by a reference program.

The tables are read from files shipped with the package
(`rx_validation_results()`). They are copies of the results of the
validation campaign held in the repository
[remlax-validation](https://github.com/NicoSalas34/remlax-validation), which
contains the scripts that produce them. Part of that campaign needs an asreml
licence and a computing cluster, so it cannot run when the package is built.
The last section shows how to rerun the comparisons against lme4, nlme and
sommer on your own machine with `rx_validate()`.

Campaign of `r vers$date`, remlax revision `r vers$remlax_revision`.
Reference software: `r paste(names(vers$versions), unlist(vers$versions), collapse = ", ")`.

## What agreement means

The restricted log-likelihood at the optimum must agree to `1e-6` absolute
(`1e-5` against asreml, which stops on its own criterion), variance
components to `1e-4` relative (`1e-3` against asreml), fixed effects to `1e-6`
relative, BLUPs to `1e-4` absolute and standard errors to `1e-4` relative.
Iteration counts are never compared.

Each program reports the log-likelihood with its own constant. lme4 and nlme
carry the full REML constant and equal `fit$logLik`. asreml omits
`(n - p)/2 log(2 pi)` and equals `fit$logLik_asreml`. sommer rescales the
response internally, so its value carries a data-dependent constant and only
its estimates, or differences of log-likelihood between two models, are
compared.

## The checks, by reference

```{r by-reference}
checks$group <- sub(" .*", "", checks$reference)
tab <- aggregate(verdict ~ group, data = checks,
                 FUN = function(v) c(checks = length(v), passed = sum(v == "pass"),
                                     failed = sum(v != "pass")))
tab <- data.frame(reference = tab$group, tab$verdict)
tab[order(-tab$checks), ]
```

`r nrow(checks)` checks, `r sum(checks$verdict == "pass")` passed. The
references called `internal`, `dense`, `independent` and `properties` are
closed forms, the sparse engine of the package compared with its dense
engine at a common parameter, a REML written longhand in dense algebra, and
properties that have no external reference (finite likelihood, positive
definite covariance).

## Largest gaps

For each program, the largest gap between remlax and the reference over all
compared quantities: absolute on the log-likelihood, relative on the
estimates. The three models discussed in the next section are shown apart.

```{r gaps}
pairs$remlax <- as.numeric(pairs$remlax); pairs$reference <- as.numeric(pairs$reference)
ecart <- ifelse(pairs$kind == "logLik", abs(pairs$remlax - pairs$reference),
                abs(pairs$remlax - pairs$reference) / pmax(abs(pairs$reference), 1e-8))
a_part <- grepl("^(ma2|sph|cir) ", pairs$check)
g <- aggregate(ecart[!a_part], by = list(software = pairs$software[!a_part],
                                          quantity = pairs$kind[!a_part]),
               FUN = function(x) signif(max(x, na.rm = TRUE), 2))
names(g)[3] <- "largest gap"
g$n <- aggregate(ecart[!a_part], by = list(pairs$software[!a_part], pairs$kind[!a_part]), FUN = length)$x
g[order(g$software, g$quantity), ]
```

```{r figure, fig.cap = "Estimates of remlax against the reference program, one point per compared quantity (log scales, absolute values). The three models discussed below are drawn as open circles."}
est <- pairs[pairs$kind == "estimate" & is.finite(pairs$remlax) & is.finite(pairs$reference) &
             abs(pairs$reference) > 1e-6 & abs(pairs$remlax) > 1e-6, ]
cols <- c(asreml = "#7f7f7f", lme4 = "#1f77b4", nlme = "#2ca02c", sommer = "#d62728")
part <- grepl("^(ma2|sph|cir) ", est$check)
op <- par(mar = c(4, 4, 1, 1))
plot(abs(est$reference), abs(est$remlax), log = "xy", pch = ifelse(part, 1, 19), cex = 0.7,
     col = cols[est$software], xlab = "|reference estimate|", ylab = "|remlax estimate|")
abline(0, 1, lty = 2, col = "grey40")
legend("topleft", legend = names(cols), col = cols, pch = 19, bty = "n", cex = 0.85)
par(op)
```

## The checks that fail, and why

```{r fails}
f <- checks[checks$verdict != "pass", c("section", "check", "detail")]
f$section <- sub(" :.*", "", f$section)
f
```

All `r nrow(f)` failures are in the asreml suite and concern three models,
the moving-average residual `ma2`, and the spherical (`sph`) and circular
(`cir`) kernels on an irregular two-dimensional field. They are left as
failures on purpose: the tolerance was not widened to make them pass.

A diagnostic compared each model with an independent REML written in dense
algebra, and held asreml at remlax's parameters.

- `sph`: asreml reports non-convergence and oscillates between two values.
  Held at remlax's parameters it returns remlax's log-likelihood to the last
  digit. Same likelihood; asreml did not converge.
- `cir`: asreml reports non-convergence and stops in the first of three local
  maxima of the likelihood in the range (ranges 5.0, 7.4 and 10.9). remlax
  reaches the highest one, 3.03 log-likelihood units above. Held at remlax's
  parameters, asreml returns remlax's value.
- `ma2`: asreml declares convergence while its parameters still move by 3.5 %
  per iteration, 0.033 below remlax. Its `cor1`/`cor2` components are not the
  moving-average coefficients of the standard form, so the parameters cannot
  be compared directly; the dense REML reproduces remlax's value exactly.

In every case remlax's log-likelihood is the higher one, and an independent
computation confirms it.

## Speed at equal model

The benchmark fitted ten models, from 2 to 157 variance parameters, at sizes
from 500 to 16000 observations, with asreml on 4 CPU cores, remlax on the same
4 cores and remlax on one NVIDIA A100. Wall time in seconds of the complete
call, for the size axis (4 replicates per genotype, 2 for the multi-trait
models):

```{r speed}
b <- bench[bench$status == "ok" & ((bench$reps == 4 & !grepl("^us", bench$case)) |
                                  (bench$reps == 2 & grepl("^us", bench$case))), ]
w <- reshape(b[, c("case", "n_nom", "serie", "wall_s")], idvar = c("case", "n_nom"),
             timevar = "serie", direction = "wide")
names(w) <- sub("wall_s.", "", names(w), fixed = TRUE)
w <- w[order(match(w$case, unique(bench$case)), w$n_nom), ]
w[, -(1:2)] <- lapply(w[, -(1:2)], function(x) signif(x, 3))
rownames(w) <- NULL
w
```

On classical designs asreml hardly depends on n: it solves sparse mixed
model equations whose size is the number of effects. remlax factorises the
dense `n x n` covariance matrix at every evaluation, so its cost grows as
`n^3`, and it is slower than asreml there. Where the problem is dense for
asreml, many genotypes for few replicates, a dense genomic relationship
matrix, weighted neighbourhood incidences, asreml's equations fill in, and
remlax on a GPU becomes faster. On the neighbourhood model (`ige`, a
single-species reduction of the model of Salas et al. 2026) the A100 is
ahead from n = 2000 and about 7 to 8 times faster at n = 8000 and 16000, at
the same log-likelihood to `1e-6`. The complete benchmark, with compilation
time, time per iteration and per evaluation, and a density axis at fixed n,
is in the remlax-validation repository.

In the benchmark no remlax fit has a lower log-likelihood than the asreml
fit it is paired with. Where remlax is higher by more than `1e-3` (five small,
heavily parameterised cases), asreml continued with `update()` moves toward
remlax's value and never passes it.

## Rerun the comparisons on your machine

`rx_validate()` refits a subset of the campaign against the reference
packages available on CRAN (lme4, nlme, sommer) and returns one row per
check. It needs the Python solver (see `rx_python_check()`), so it is not run
when this vignette is built without Python.

```{r rerun, eval = a_python}
v <- rx_validate(verbose = FALSE)
v[, c("reference", "case", "quantity", "gap", "tolerance", "pass")]
attr(v, "versions")
```

To rerun the full campaign, including asreml and the benchmark, clone
remlax-validation and follow its README.

Salas N, Montazeaud G, Bourke PM, Baranger A, David J (2026). Multispecies
mixtures: an individual-centered quantitative genetic framework for complex
plant neighborhoods. bioRxiv. doi:10.64898/2026.05.27.728303.